Supporting Information:\n prolfqua - A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-10-30 inst/MyArticle/paper.Rmd
added test for difference between MaxQuant and FragPipe

---
title: prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [University of Zurich](https://uzh.ch)/[ETH Zurich](https://ethz.ch), Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland](https://www.sib.swiss/) $^\dagger$Correspondence&colon; \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
      citation_package: natbib
date: "`r format(Sys.time())`"
bibliography: prolfqua.bib
vignette: >
  %\VignetteIndexEntry{prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. There is a large variety of signal processing software and analysis tools. Nevertheless, there is a need for a modular, easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures to make applying them to proteomics data, comparing and understanding their differences easily. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The package makes integrating new data formats easy. It can be used to model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. The implemented methods allow sensitive and specific differential expression analysis. Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. The application programmer interface of `r BiocStyle::Githubpkg("fgcz/prolfqua")`  strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis application development easy and exciting. Finally, the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT license. It runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).

@@

The package *prolfqua* provides functions to determine the vector of $parameter$ weights $c$, from a linear model and a contrast specification string. In the Supplement we provided an example, how to specify contrasts for a dataset with two explanatory variable with an interaction term (Supplementary Section "Specifying Contrasts for Models with two Factors and Interaction Term").

@@

In addition, we examined the DEA performance when using protein intensities reported by quantification software *FragPipe* for the *IonStar* dataset as input. Using protein abundances as input significantly simplifies the analysis and interpretation and might benefit from optimization implemented in the signal processing software. However, we can only easily fit the *proDA* and `prolfqua_merged` models to protein intensities. In this DEA benchmark, *prolfqua* performed slightly, but not significantly better than *proDA*  (Supplementary Section "DEA benchmark : IonStar/FragPipeV14/combined_protein.tsv").

@@

We focused our benchmark on comparing the statistical modeling methods while we fixed the pre-processing steps. However, some of these steps are of utmost significance when performing differential expression analysis [@frohlich2022benchmarking]. One of them is the normalization of the intensities within the samples to remove systematic differences [@pursiheimo2015optimization]. The method used to infer proteins from peptide identifications [@nesvizhskii2005interpretation] and protein abundances from peptide abundances is an additional important factor [@Grossmann2010]. For instance, the original `r Biocpkg("proDA")` publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ intensities reported by *MaxQuant*, the $pAUC_{10}$ is lower ($pAUC_{10}(\textrm{t-statistics}) = 66%$) compared with results obtained when protein abundances are estimated from peptides abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 72%$). Last but not least, the software [@Cox2008MaxQuant; @yu2020fast] used to identify and quantify proteins significantly contributes to the entire pipeline's performance altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis. In the Supplementary Section "Comparing DEA results for *MaxQuant* and *FragPipe* we compare DEA benchmarking results for *MaxQuant* and *FragPipe*. While the number of proteins identified with two peptides is practically the same, the DEA benchmark performance differs significantly by $\approx 10%$ for the $pAUC_{10}$ score.

@@

The authors thank the technology platform fund (TPF) of the University of Zurich and all colleagues of the FGCZ proteomics team for fruitful discussions.

@@

\begin{table}
\centering
\begin{tabular}{ll}
\toprule
Abbreviations & Explanation\\
\midrule
API    & application programming interface          \\
AUC    & Area Under the Curve                               \\
CDF    & Cumulative Distribution Function                   \\
DIA    & data independent acquisition                        \\
ESI-MS & Electro-Spray-Ionization Mass Spectrometry         \\
FAIR   & Findable, Accessible, Interoperable, and Reusable  \\
$FDP$  & False Discovery Proportion                         \\
$FDR$  & False Discovery Rate                               \\
LC     & Liquid Chromatography                              \\
LC-MS  & Liquid Chromatography followed by Mass Spectrometry\\
LOD    & Limit Of Detection                                 \\
MAR    & Missing At Random                                 \\ 
MCAR   & Missing Completely At Random                      \\
ML     & machine learning                                 \\
MS     & mass spectrometry                                  \\
pAUC   & partial Are Under the Curve                       \\
ROC    & Receiver operator curve                           \\
TMT    & Tandem Mass Tag                                   \\
UML    & Unified Modeling Language                        \\
\bottomrule
\end{tabular}
\end{table}

>>>>

---
title: prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [University of Zurich](https://uzh.ch)/[ETH Zurich](https://ethz.ch), Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland](https://www.sib.swiss/) $^\dagger$Correspondence&colon; \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
      citation_package: natbib
date: "`r format(Sys.time())`"
bibliography: prolfqua.bib
vignette: >
  %\VignetteIndexEntry{prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. There is a large variety of quantification software and analysis tools. Nevertheless, there is a need for a modular, easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures to make applying them to proteomics data, comparing and understanding their differences easily. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The package makes integrating new data formats easy. It can be used to model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. The implemented methods allow sensitive and specific differential expression analysis. Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. The application programmer interface of `r BiocStyle::Githubpkg("fgcz/prolfqua")`  strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis application development easy and exciting. Finally, the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT license. It runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).

@@

The package *prolfqua* provides functions to determine the vector of $parameter$ weights $c$, from a linear model and a contrast specification string. In the Supplement we provided an example, how to specify contrasts for a dataset with two explanatory variables and an interaction term (Supplementary Section "Specifying Contrasts for Models with two Factors and Interaction Term").

@@

In addition, we examined the DEA performance when using protein intensities reported by quantification software *FragPipe* for the *IonStar* dataset as input. Using protein abundances as input significantly simplifies the analysis and interpretation and might benefit from optimization implemented in the quantification software. However, we can only easily fit the *proDA* and `prolfqua_merged` models to protein intensities. In this DEA benchmark, *prolfqua* performed slightly, but not significantly better than *proDA*  (Supplementary Section "DEA benchmark : IonStar/FragPipeV14/combined_protein.tsv").

@@

We focused our benchmark on comparing the statistical modeling methods while we fixed the pre-processing steps. However, some of these steps are of utmost significance when performing differential expression analysis [@frohlich2022benchmarking]. One of them is the normalization of the intensities within the samples to remove systematic differences [@pursiheimo2015optimization]. The method used to infer proteins from peptide identifications [@nesvizhskii2005interpretation] and protein abundances from peptide abundances is an additional important factor [@Grossmann2010]. For instance, the original `r Biocpkg("proDA")` publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ intensities reported by *MaxQuant*, the $pAUC_{10}$ is lower ($pAUC_{10}(\textrm{t-statistics}) = 66%$) compared with results obtained when protein abundances are estimated from peptides abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 72%$). Last but not least, the software [@Cox2008MaxQuant; @yu2020fast] used to identify and quantify proteins significantly contributes to the entire pipeline's performance altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis. In the Supplementary Section "Comparing DEA results for *MaxQuant* and *FragPipe* we compare DEA benchmarking results for *MaxQuant* and *FragPipe*. While the number of proteins identified with two peptides is practically the same, the DEA benchmark performance differs significantly by $\approx 10%$ for the $pAUC_{10}$ score. This difference is more significant than the differences due to the choice of the modeling method.

@@

The authors thank the technology platform fund of the University of Zurich and all colleagues of the FGCZ proteomics team for fruitful discussions.

@@

\begin{table}
\centering
\begin{tabular}{ll}
\toprule
Abbreviations & Explanation\\
\midrule
API    & application programming interface          \\
AUC    & Area Under the Curve                               \\
CDF    & Cumulative Distribution Function                   \\
DEA & Differential Expression Analysis \\
DIA    & data independent acquisition                        \\
ESI-MS & Electro-Spray-Ionization Mass Spectrometry         \\
FAIR   & Findable, Accessible, Interoperable, and Reusable  \\
$FDP$  & False Discovery Proportion                         \\
$FDR$  & False Discovery Rate                               \\
$FP$ & False Postive \\
$FPR$ & False Positive Rage    \\
LC     & Liquid Chromatography                              \\
LC-MS  & Liquid Chromatography followed by Mass Spectrometry\\
LOD    & Limit Of Detection                                 \\
MAR    & Missing At Random                                 \\ 
MCAR   & Missing Completely At Random                      \\
ML     & machine learning                                 \\
MS     & mass spectrometry                                  \\
pAUC   & partial Area Under the Curve                       \\
$pAUC_{10}$ & partial Area Unter the Curve for an FPR of 0 to 10% \\
ROC    & Receiver operator curve                           \\
TP & True Positive \\
TMT    & Tandem Mass Tag                                   \\
UML    & Unified Modeling Language                        \\
\bottomrule
\end{tabular}
\end{table}

### 2022-10-30 inst/MyArticle/prolfqua_supplement.Rmd
added test for difference between MaxQuant and FragPipe

```{r shwoROCtestREsultsIonStarMQ}
cap = "p-values for pairwise comparsions of partial AUC at 0.1 for msqrob2, proDA and prolfqua for IonStar MaxQuant data."
knitr::kable(tibble::tibble(pROC_test = names(tmp), p.value = round(tmp, digits = 2)),
             format = "latex", caption = cap)

@@

```{r shwoROCtestREsultsCPTAC}
cap = "p-values for pairwise comparsions of partial AUC for msqrob2, proDA and prolfqua CPTAC/MaxQuant data."
knitr::kable(tibble::tibble(pROC_test = names(tmp), p.value = round(tmp, digits = 2)),
             format = "latex",
             caption = cap)

@@

```{r shwoROCtestREsultsFRAGPIPE}
cap = "p-values for pairwise comparsions of partial AUC for proDA and prolfqua CPTAC/MaxQuant data."
knitr::kable(tibble::tibble(pROC_test = names(tmp), p.value = round(tmp, digits = 2)),
             format = "latex",
             caption = cap)

@@

In the Article, we discussed other factors influencing the performance of the DEA analysis. One of these factors is the quantification software. Here we compare the benchmarking results obtained with MaxQuant and FragPipe. We see (Supplementary Figure \@ref(fig:compareMaxQuantFP)) that the choice of the signal processing software changes the partial area under the curve ($pAUC_{10}$) much stronger (by up to $8%$) than the DEA analysis method (by about 4%).
Furthermore, we observe that the DEA benchmark results differ depending on if _MSstats.tsv_ or _combined_protein.tsv_ is used as input.

(ref:compareMaxQuantFP) Barplot, showing the DEA results ($pAUC_{10}$ - y axis) for four DEA methods and two signal processing methods (FragPipeV14/MSstats.tsv in black, FragPipeV14/combined_protein.txt dark gray and MaxQuant/peptide.txt in gray).

@@

>>>>

```{r shwoROCtestREsultsIonStarMQ}
cap = "p-values for pairwise comparsions of $pAUC_{10}$ for msqrob2, proDA and prolfqua for IonStar MaxQuant data."
knitr::kable(tibble::tibble(pROC_test = names(tmp), p.value = round(tmp, digits = 2)),
             format = "latex", caption = cap)

@@

```{r shwoROCtestREsultsCPTAC}
cap = "p-values for pairwise comparsions of $pAUC_{10}$ for *msqrob2*, *proDA* and *prolfqua* CPTAC/MaxQuant/peptide.txt data."
knitr::kable(tibble::tibble(pROC_test = names(tmp), p.value = round(tmp, digits = 2)),
             format = "latex",
             caption = cap)

@@

```{r shwoROCtestREsultsFRAGPIPE}
cap = "p-values for pairwise comparsions of $pAUC_{10}$ for proDA and prolfqua CPTAC/MaxQuant data."
knitr::kable(tibble::tibble(pROC_test = names(tmp), p.value = round(tmp, digits = 2)),
             format = "latex",
             caption = cap)

@@

In the Article, we discussed other factors influencing the performance of the DEA analysis. One of these factors is the quantification software. Here we compare the benchmarking results obtained with MaxQuant and FragPipe. We see (Supplementary Figure \@ref(fig:compareMaxQuantFP)) that the choice of the quantification software changes the partial area under the curve ($pAUC_{10}$) much stronger (by up to $8%$) than the DEA analysis method (by about 4%).
Furthermore, we observe that the DEA benchmark results differ depending on if _MSstats.tsv_ or _combined_protein.tsv_ is used as input (see Figure \@ref(fig:compareMaxQuantFP) Left Panel, methods *proDA* and *prolfqua*). 
Supplementary Table \@ref(tab:shwoROCtestMQandFP) shows the p-values when we test if the $pAUC_{10}$, summarizing the DEA benchmark results, significantly differ between the *MaxQuant* and *FragPipe*. At significance level of $1%$ we will reject the null hypothesis that there is no difference between *FragPipe* and *MaxQuant*.

(ref:compareMaxQuantFP) Barplot, showing the DEA results ($pAUC_{10}$ - y axis) for four DEA methods and two quantification methods (FragPipeV14/MSstats.tsv in black, FragPipeV14/combined_protein.txt dark gray and MaxQuant/peptide.txt in gray).

@@

```{r roctestFPMaxQuant, include = TRUE, eval = TRUE}
allBenchmarks <- readRDS("allBenchmarks.RDS")
allB <- readRDS(getpath("FragPipev14_comb_MSStats.RDS"))

ttprolfquaMQ <- allBenchmarks$benchmark_merged$data()
ddprolfquaMQp.val <- pROC::roc(ttprolfquaMQ$TP, ttprolfquaMQ$scaled.p.value, partial.auc = c(1, 0.9))
ddprolfquaMQdiff <- pROC::roc(ttprolfquaMQ$TP, ttprolfquaMQ$diff, partial.auc = c(1, 0.9))

ttprolfquaFP <- allB$benchmark_prolfqua$data()
ddprolfquaFPp.val <- pROC::roc(ttprolfquaFP$TP, ttprolfquaFP$scaled.p.value, partial.auc = c(1, 0.9))
ddprolfquaFPdiff <- pROC::roc(ttprolfquaFP$TP, ttprolfquaFP$diff, partial.auc = c(1, 0.9))

tmp <- c(MQ_vs_FP_scaled.p.value = pROC::roc.test(ddprolfquaMQp.val,ddprolfquaFPp.val, progress = "none")$p.value,
         MQ_vs_FP_diff = pROC::roc.test(ddprolfquaMQdiff, ddprolfquaFPdiff, progress = "none")$p.value)
tmp
```

```{r shwoROCtestMQandFP}
cap = "p-values for pairwise comparsions of $pAUC_{10}$ for MaxQuant and FragPipe based DEA analysis with prolfqua. We compared the pAUC for the difference estimates (diff) and for the scaled.p.values."
knitr::kable(tibble::tibble(pROC_test = names(tmp), p.value = round(tmp, digits = 2)),
             format = "latex", caption = cap)
